Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-29 draft.tex
finished related work section, still has to be revised

Epidemiologists want to characterize the healthy aging process of the spine.
%%
To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
%% -------------------------------
\section{\uppercase{Related Work}}
\label{sec:RelatedWork}
\noindent \com{Our own work (VIS, VMV, BVM). Sylvias paper with reference to the methods. \textbf{More information necessary here!}}

\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

>>>>

Epidemiologists want to characterize the healthy aging process of the spine.
%%
To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.
%% -------------------------------
\section{\uppercase{Related Work}}
\label{sec:RelatedWork}
%%
\paragraph{Visual Analysis of Image and non-Image Data}
%%
Klemm et al. \cite{Klemm2014VIS} propose an extension to the existing epidemiological workflow using Interactive Visual Analysis (IVA).
%%
The workflow consists of an iterative sequence of group selection, variable selection and visualization.
%%
It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.
%%
We incorporate the IVA approach by concurrently displaying complex relationships between variables using decision trees.
%%
Steenwijk et al. \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
%%
It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots with brushing and linking facilities to successively analyze a data set.
%%
Similar to our work, they quantify image data to map them into the same information space as non-image data.
%%
Turkay et al. \cite{Turkay} follow a similar approach by deriving descriptive data metrics from image data.
%%
Their proposed \emph{deviation plot} shows distribution-specific measures of a variable, such as skewness or inter-quantile-range, making dimensions comparable.
%%
This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these dimensions.

\paragraph{Decision Rule Driven Analysis of Medical Data.}
Closest to our work is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}.
%%
Related work in this field often focuses on clinical diagnosis.
%%
Gla{\ss}er et al. \cite{Glasser2013} use parameters derived from DCE-MRI image data to classify breast tumors.
%%
Their parameters are derived from clustering of bloodflow in the tumorous tissue.
%%
These features trains a decision-tree classifier, concluding that the extracted features alone are not sufficient for tumor type classification.
%%
Niemann et al. \cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.
%%
They presented an interactive data mining tool, which can analyze association rules and highlight interesting relations, which may trigger new hypothesis about the data.
%%
The presented paper either validate the significance of image derived parameter or derive new relationships by training decision trees to explain target variables.
%%

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and understandable and are therefore well suited for classifying medical data.
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, however, get too complex, analyzing these decision rules can get complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization techniques for decision trees, where all decisions are displayed in a heat map
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing a decision tree, but not for multiple trees.
%%
\paragraph{Visual Analysis of Heterogenous Non-Image Data.}
%%
Zhang et al. \cite{Zhang} analyze subject groups in a web-based linked view system.
%%
The resulting decision rules aim to categorize new subjects as they are added to the data.
%%
They define a cohort as parameter-divided subject group, differing from the epidemiological understanding of a cohort.
%%
The paper lacks details about handling of missing data, the definition of similarity or which statistical measures are used.
%%
Generalized Pairs Plots (\texttt{GPLOM'S}) visualize heterogenous variables in a plot matrix \cite{GPLOMS, Francois2013}.
%%
The plot depends on the type of variables, which are pairwise visualized.
%%
The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{image-parameter-range}).
%%
Dai et al. \cite{Dai} defined a \emph{Concept Map} linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
%%
The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.
%%
Our work does not contain geographical information of any kind.
%%----------------------------
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay} to make them comparable in a overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%
